Geometric and physical interpretation of the action principle

### 2022-10-13 Articles/2022-ActionPrinciple/ActionPrincipleArchive.tex
ActionPrinciple: adding note on nonholonomic constraints
ActionPrinciple: geometrical -> geometric

\title{Geometrical and physical interpretation of the action principle}
\author{Gabriele Carcassi, Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
We give a geometrical interpretation for the principle of stationary action in classical Lagrangian particle mechanics. In a nutshell, the difference of the action along a path and its variation effectively ``counts'' the possible evolutions that ``go through'' the area enclosed. If the path corresponds to a possible evolution, all neighbouring evolutions will be parallel, making them tangent to the area enclosed by the path and its variation, thus yielding a stationary action. This treatment gives a full physical account of the geometry of both Hamiltonian and Lagrangian mechanics which is founded on three assumptions: determinism and reversible evolution, independence of the degrees of freedom and equivalence between kinematics and dynamics. The logical equivalence between the three assumptions and the principle of stationary action leads to a much cleaner conceptual understanding.
\end{abstract}

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometrical interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometrical meaning, leaving the mathematical details and calculations to the appendix.

@@

We have thus demystified the action principle, and turned it into a geometrical property: requiring the principle of stationary action is equivalent to requiring that the solutions are the field lines of a divergence-free field in phase space. We also have a clear physical meaning: the principle of stationary action is equivalent to assuming determinism/reversibility \ref{assum_detrev} and kinematic equivalence \ref{assum_kineq}. However, we do feel that the principle expresses these requirements in a very roundabout way.

@@

The above is the generalization of \ref{sdof_displacement_potential} and it is important to note the differences. The exterior derivative takes the place of the curl and it is $\omega$, not $\vec{S}$, that can be expressed in terms of the potential $\theta$. Yet, there is an important geometrical relationship we still need to take into account: the direction of the displacement does not contribute new states. Physically, this is a consequence of \ref{assum_detrev}: states cannot be created or disappear over time. We have
\begin{align}\label{mdof_displacement_kills}
	\omega(\vec{S}, \cdot) = 0.
\end{align}
Mathematically we say that the displacement vector kills the form; it makes it zero no matter what vector we put on the other side.

@@

We have found that the principle of stationary action is equivalent to assumptions \ref{assum_detrev}, \ref{assum_indep} and \ref{assum_kineq} and that the variation of the action can be understood as the flow of evolutions through the surface delimited by the path and its variation. This allows us to fully understand both the physical and geometrical significance of the principle in a much more precise way.

@@

As a final comment, note that privileging the Hamiltonian picture is more in line with quantum mechanics; the idea of $\omega$ as a tool to count states is in line with statistical mechanics; expression \ref{mdof_potential_expression} is remarkably similar to that of the relativistic four-momentum. This is one of the key insights we get from our Reverse Physics approach: there is a unity among the different physical theories that begs to be brought to light. We are convinced that a version of this geometrical understanding must exist in the quantum world.

@@

For (iii), we expand each component of \ref{sdof_displacement_potential} using \ref{sdof_potential_expression}. Again note the double role of time: $\frac{dp}{dt}$ is not the same as $\frac{\partial p}{\partial t}$.\footnote{We could be more precise and use $\tau$ for the evolution parameter, leaving $t$ just for the coordinate. The treatment would automatically acquire a relativistic flavor and would enlighten us to other interesting results, though it would distract from the main focus of this paper, the action principle. } In the first case we are taking a total derivative along the evolution, and therefore the momentum can change. In the second case we are taking a partial derivative at constant $q$ and $p$ by definition, and therefore $\frac{\partial p}{\partial t}=0$. We have
\begin{equation}
\begin{aligned}
	S^q = \frac{dq}{dt}
	&= - \left( \frac{\partial}{\partial p} \theta_t - \frac{\partial}{\partial t} \theta_p \right) \\
	&= - \left( \frac{\partial}{\partial p} (-H) - \frac{\partial}{\partial t} 0 \right) \\
	& = \frac{\partial H}{\partial p}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
	S^p = \frac{dp}{dt}
	&= - \left( \frac{\partial}{\partial t} \theta_q - \frac{\partial}{\partial q} \theta_t \right) \\
	&= - \left( \frac{\partial}{\partial t} p - \frac{\partial}{\partial q} (-H) \right) \\
	& = - \frac{\partial H}{\partial q}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
	S^t = \frac{dt}{dt}
	&= - \left( \frac{\partial}{\partial q} \theta_p - \frac{\partial}{\partial p} \theta_q \right) \\
	&= - \left( \frac{\partial}{\partial q} 0 - \frac{\partial}{\partial p} p \right) \\
	& = 1
\end{aligned}
\end{equation}
We have recovered Hamilton's equations
\begin{equation}
\tag*{(\ref{sdof_Ham_eq})}
\begin{aligned}
	\frac{dq}{dt} &= \frac{\partial H}{\partial p} \\
	\frac{dp}{dt} &=-\frac{\partial H}{\partial q}
\end{aligned}
\end{equation}
as the equations for a deterministic and reversible system. The only thing that the equations say is that states move in time at the same rate (i.e. $\frac{dt}{dt} = 1$) with an incompressible flow (i.e. no states are created or destroyed). That is the whole physical and geometrical content of those equations.

@@

Now that we have developed a clear geometrical and physical understanding of Hamilton's equations, we can concentrate on the action principle. For (iv) we can apply Stokes' theorem. We take a path $\gamma$ and a variation $\gamma'$. Since they share the endpoints, taken together they form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\vec{S}$ is divergence-free, the flow of $\vec{S}$ through a surface depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Putting it all together we have
\begin{equation}
\tag*{(\ref{sdof_action})}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &=
	\iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	&= \oint_{\partial \Sigma} \vec{\theta}  \cdot d\vec{\gamma} \\
	&= \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} - \int_{\gamma'} \vec{\theta} \cdot d\vec{\gamma}' \\
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

@@

We want to stress that this variational principle applies to any divergence-free field. For example, for a magnetic field we could write:
\begin{equation}
	\iint_{\Sigma} \vec{B} \cdot d\vec{\Sigma} = \delta \int_{\gamma} \vec{A} \cdot d\vec{\gamma}.
\end{equation}
We would find that the field lines of $\vec{B}$ are the lines and only the lines for which the integral of the vector potential is stationary. Similarly, we would find that the streamlines of an incompressible fluids are the lines and only the lines for which the integral of the vector potential of the flow is stationary. While these properties may be of limited practical use, the geometrical insight is the same.

@@

What we will give here, then, is not a fully rigorous introduction to the subject, but rather a geometrical understanding of what these tools describe and a physical motivation as to why they are useful. We hope this will make it easier to follow the next section and to build a geometrical and physical intuition of the principle of stationary action, which is the focus of this article.

@@

%In our opinion, , and In order to fully understand the geometrical and physical meaning of the general case, it helps to have a crisp intuition for the tools of differential topology. We will therefore go through the basic motivation of differential topology from a physicist perspective, so that we can see what differential forms and exterior derivatives are capturing at a a physical level.

@@

Note that we have broken down exactly how each assumption contributes to the mathematical setting. If \ref{assum_indep} fails, then $\Omega$ cannot be understood as $\omega^N$, which means we have much fewer conditions on the motion: we only know that the total flow is conserved, but we cannot break it down along each d.o.f.. On the other hand, if \ref{assum_detrev} fails we may not have a single displacement field $\vec{S}$ and/or the form that counts states would not be closed. This type of exact mapping between assumptions and their mathematical expression is one of the key goals of Reverse Physics in particular and our overall project Assumptions of Physics more in general.

>>>>

\title{Geometric and physical interpretation of the action principle}
\author{Gabriele Carcassi, Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
We give a geometric interpretation for the principle of stationary action in classical Lagrangian particle mechanics. In a nutshell, the difference of the action along a path and its variation effectively ``counts'' the possible evolutions that ``go through'' the area enclosed. If the path corresponds to a possible evolution, all neighbouring evolutions will be parallel, making them tangent to the area enclosed by the path and its variation, thus yielding a stationary action. This treatment gives a full physical account of the geometry of both Hamiltonian and Lagrangian mechanics which is founded on three assumptions: determinism and reversible evolution, independence of the degrees of freedom and equivalence between kinematics and dynamics. The logical equivalence between the three assumptions and the principle of stationary action leads to a much cleaner conceptual understanding.
\end{abstract}

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the appendix.

@@

We have thus demystified the action principle, and turned it into a geometric property: requiring the principle of stationary action is equivalent to requiring that the solutions are the field lines of a divergence-free field in phase space. We also have a clear physical meaning: the principle of stationary action is equivalent to assuming determinism/reversibility \ref{assum_detrev} and kinematic equivalence \ref{assum_kineq}. However, we do feel that the principle expresses these requirements in a very roundabout way.

@@

The above is the generalization of \ref{sdof_displacement_potential} and it is important to note the differences. The exterior derivative takes the place of the curl and it is $\omega$, not $\vec{S}$, that can be expressed in terms of the potential $\theta$. Yet, there is an important geometric relationship we still need to take into account: the direction of the displacement does not contribute new states. Physically, this is a consequence of \ref{assum_detrev}: states cannot be created or disappear over time. We have
\begin{align}\label{mdof_displacement_kills}
	\omega(\vec{S}, \cdot) = 0.
\end{align}
Mathematically we say that the displacement vector kills the form; it makes it zero no matter what vector we put on the other side.

@@

We have found that the principle of stationary action is equivalent to assumptions \ref{assum_detrev}, \ref{assum_indep} and \ref{assum_kineq} and that the variation of the action can be understood as the flow of evolutions through the surface delimited by the path and its variation. This allows us to fully understand both the physical and geometric significance of the principle in a much more precise way.

@@

As a final comment, note that privileging the Hamiltonian picture is more in line with quantum mechanics; the idea of $\omega$ as a tool to count states is in line with statistical mechanics; expression \ref{mdof_potential_expression} is remarkably similar to that of the relativistic four-momentum. This is one of the key insights we get from our Reverse Physics approach: there is a unity among the different physical theories that begs to be brought to light. We are convinced that a version of this geometric understanding must exist in the quantum world.

@@

For (iii), we expand each component of \ref{sdof_displacement_potential} using \ref{sdof_potential_expression}. Again note the double role of time: $\frac{dp}{dt}$ is not the same as $\frac{\partial p}{\partial t}$.\footnote{We could be more precise and use $\tau$ for the evolution parameter, leaving $t$ just for the coordinate. The treatment would automatically acquire a relativistic flavor and would enlighten us to other interesting results, though it would distract from the main focus of this paper, the action principle. } In the first case we are taking a total derivative along the evolution, and therefore the momentum can change. In the second case we are taking a partial derivative at constant $q$ and $p$ by definition, and therefore $\frac{\partial p}{\partial t}=0$. We have
\begin{equation}
\begin{aligned}
	S^q = \frac{dq}{dt}
	&= - \left( \frac{\partial}{\partial p} \theta_t - \frac{\partial}{\partial t} \theta_p \right) \\
	&= - \left( \frac{\partial}{\partial p} (-H) - \frac{\partial}{\partial t} 0 \right) \\
	& = \frac{\partial H}{\partial p}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
	S^p = \frac{dp}{dt}
	&= - \left( \frac{\partial}{\partial t} \theta_q - \frac{\partial}{\partial q} \theta_t \right) \\
	&= - \left( \frac{\partial}{\partial t} p - \frac{\partial}{\partial q} (-H) \right) \\
	& = - \frac{\partial H}{\partial q}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
	S^t = \frac{dt}{dt}
	&= - \left( \frac{\partial}{\partial q} \theta_p - \frac{\partial}{\partial p} \theta_q \right) \\
	&= - \left( \frac{\partial}{\partial q} 0 - \frac{\partial}{\partial p} p \right) \\
	& = 1
\end{aligned}
\end{equation}
We have recovered Hamilton's equations
\begin{equation}
\tag*{(\ref{sdof_Ham_eq})}
\begin{aligned}
	\frac{dq}{dt} &= \frac{\partial H}{\partial p} \\
	\frac{dp}{dt} &=-\frac{\partial H}{\partial q}
\end{aligned}
\end{equation}
as the equations for a deterministic and reversible system. The only thing that the equations say is that states move in time at the same rate (i.e. $\frac{dt}{dt} = 1$) with an incompressible flow (i.e. no states are created or destroyed). That is the whole physical and geometric content of those equations.

@@

Now that we have developed a clear geometric and physical understanding of Hamilton's equations, we can concentrate on the action principle. For (iv) we can apply Stokes' theorem. We take a path $\gamma$ and a variation $\gamma'$. Since they share the endpoints, taken together they form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\vec{S}$ is divergence-free, the flow of $\vec{S}$ through a surface depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Putting it all together we have
\begin{equation}
\tag*{(\ref{sdof_action})}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &=
	\iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	&= \oint_{\partial \Sigma} \vec{\theta}  \cdot d\vec{\gamma} \\
	&= \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} - \int_{\gamma'} \vec{\theta} \cdot d\vec{\gamma}' \\
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

@@

We want to stress that this variational principle applies to any divergence-free field. For example, for a magnetic field we could write:
\begin{equation}
	\iint_{\Sigma} \vec{B} \cdot d\vec{\Sigma} = \delta \int_{\gamma} \vec{A} \cdot d\vec{\gamma}.
\end{equation}
We would find that the field lines of $\vec{B}$ are the lines and only the lines for which the integral of the vector potential is stationary. Similarly, we would find that the streamlines of an incompressible fluids are the lines and only the lines for which the integral of the vector potential of the flow is stationary. While these properties may be of limited practical use, the geometric insight is the same.

@@

What we will give here, then, is not a fully rigorous introduction to the subject, but rather a geometric understanding of what these tools describe and a physical motivation as to why they are useful. We hope this will make it easier to follow the next section and to build a geometric and physical intuition of the principle of stationary action, which is the focus of this article.

@@

%In our opinion, , and In order to fully understand the geometric and physical meaning of the general case, it helps to have a crisp intuition for the tools of differential topology. We will therefore go through the basic motivation of differential topology from a physicist perspective, so that we can see what differential forms and exterior derivatives are capturing at a a physical level.

@@

Note that we have broken down exactly how each assumption contributes to the mathematical setting. If \ref{assum_indep} fails, then $\Omega$ cannot be understood as $\omega^N$, which means we have much fewer conditions on the motion: we only know that the total flow is conserved, but we cannot break it down along each d.o.f.. On the other hand, if \ref{assum_detrev} fails we may not have a single displacement field $\vec{S}$ and/or the form that counts states would not be closed. Nonholonomic constraints will, in general, make one of the assumptions fail, for example by introducing a dependence between degrees of freedom. This type of exact mapping between assumptions and their mathematical expression is one of the key goals of Reverse Physics in particular and our overall project Assumptions of Physics more in general.
